The Galactic Midplane is Not a Plane: Implications for dynamical analysis with {\em Gaia} data and beyond

### 2019-04-22 paper/ms.tex
Small typo fixes from MKN

Stellar actions, computed for stars from their six-dimensional phase space measurement and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. 
%mm Of course,
    Such analyses are often used to interpret data from the {\em Gaia}
    mission. However, 
inaccurate measurements of the 
phase space position and uncertainties in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as 
%mm an 
    a systematic
error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on
%mm  the true action value. [the error in a Gaussian peaks on zero] 
    the null value.
A variation in the vertical position midplane of $\sim15\,\pc$ for a thin disk orbit or
$\sim120\,\pc$ for a thick disk orbit is 
%mm all that is necessary 
      sufficient
to induce a $25\,\%$ systematic error in the vertical action $J_z$ computed under the assumption of an axisymmetric potential.
%mm [fragment?] orbits, respectively. 
Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic 
%mm offset in $z$.
   vertical offset variation. 
Variation in the local standard of rest
%mm will induce 
     induces
similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across
the large regions of the disk
accessible 
%mm with the {\em Gaia} mission. [moved first mention up to top of
%abstract, given that this is in the title.]
    to {\em Gaia}.

@@

\subsection{Local Midplane} \label{ssec:local_midplane}
Using the three simulations, we determine the local midplane that an observer
might measure if they were situated in each of these galaxies as a function of
azimuth at the solar circle. Starting from the coordinate system described in
the previous section, which is approximately aligned so that the
$z$-coordinate is perpendicular to the disk plane, we place our imaginary
observer at $z=0$ and a solar radius of $8.2\,\kpc$ and vary the azimuth
between $0<\phi<2\pi$\footnote{Our very long-lived observer also has warp
drive.}. At each value of $\phi$ we then compute the median $z$ for stars
within a cylinder of radius $0.5\,\kpc$ and height $1\,\kpc$ perpendicular to
the fiducial disk. We then re-define this median $z$ as the new midplane of
the cylinder, re-select stars, and iterate until the median $z$ value
converges. We find that only $10$ iterations of this procedure are necessary
for convergence. The resulting median $z$ is taken to be what our observer
would measure as the local Galactic midplane at each $\phi$. This procedure
assumes perfect density estimation, and therefore perfect corrections for
extinction, within the cylinder defining the ``solar neighborhood.'' Imperfect
extinction correction is likely to increase the amplitude of the estimated
fluctuations in $z$. To account for the effect of particle noise, we
bootstrap-resample stars within a cylinder of stars with height $2\,\kpc$ and
the same radius $1000$ times and determine the $1\,\sigma$ error bars. This
bootstrap resampling is performed by first selecting all stars within a height
of $2\,\kpc$, resampling that selection, and then repeating the $10$
iterations.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=7in]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels). The local midplane is determined at a
position $\phi$ by taking the median height of all stars within $R=0.5\,\kpc$
and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma
$error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

@@

\begin{figure*}
\plotone{fig/range_dphi.pdf}
\caption{The range of midplane heights encountered as a function of angular
width. At each angle $\phi$ from Figure~\ref{fig:midplane} we consider an
angular width of $\Delta \phi$ centered on $\phi$ and report the range of
midplane heights within that width. We repeat the procedure for each $\phi$
and plot the result as translucent gray lines. We also plot the mean range as
a solid blue line and the $\pm1\sigma$ lines as dashed blue lines. The upper
$x$-axis shows the chord length from the position $\phi$ to $\pm\Delta\phi/2$.
The horizontal \thincolor{} and \thickcolor{} lines shows the midplane
variation necessary to induce a $10\%$ error in $J_z$ for the thin-disk and
thick-disk orbits, respectively. A cartoon explanation of this Figure is given
in Figure~\ref{fig:fig_to_explain}. We only consider the Solar circle to
explore the impact of azimuthal midplane variations, but similar effects are
expected for radial midplane variations.}
\label{fig:range_deltaphi}
\end{figure*}

@@

We also compute $R_{\n}$ for each of the orbits of the real open clusters in
Table~\ref{tab:real_clusters}, using the midplane variations from
\mi{} as a conservative choice, since it has the largest degree of
midplane variation (Figure~\ref{fig:midplane}). We report the neighborhood
values for the mass of each cluster in the neighborhood column of
Table~\ref{tab:real_clusters}. Except for alphaPer, we find that the
neigbhorhood around each cluster is typically $\gtrsim x\,\pc$. This analysis
assumes that the MW resembles \mi{}, which is not necessarily true. However,
this step of our analysis is meant to be more illustrative than quantitative:
this calculation should actually be performed using real measurements of
variations in the MW's stellar midplane. Such measurements should be possible
in the near future with {\em Gaia}, with quality distance estimates the
crucial limiting factor, along with careful selction function modelling.

>>>>

Stellar actions, computed for stars from their six-dimensional phase space measurements and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. 
%mm Of course,
    Such analyses are often used to interpret data from the {\em Gaia}
    mission. However, 
inaccurate measurements of the 
phase space position and uncertainties in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as 
%mm an 
    a systematic
error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on
%mm  the true action value. [the error in a Gaussian peaks on zero] 
    the null value.
A variation in the vertical position midplane of $\sim15\,\pc$ for a thin disk orbit or
$\sim120\,\pc$ for a thick disk orbit is 
%mm all that is necessary 
      sufficient
to induce a $25\,\%$ systematic error in the vertical action $J_z$ computed under the assumption of an axisymmetric potential.
%mm [fragment?] orbits, respectively. 
Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic 
%mm offset in $z$.
   vertical offset variation. 
Variation in the local standard of rest
%mm will induce 
     induces
similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across
the large regions of the disk
accessible 
%mm with the {\em Gaia} mission. [moved first mention up to top of
%abstract, given that this is in the title.]
    to {\em Gaia}.

@@

\subsection{The Local Midplane} \label{ssec:local_midplane}
Using the three simulations, we determine the local midplane that an observer
might measure if they were situated in each of these galaxies as a function of
azimuth at the solar circle. Starting from the coordinate system described in
the previous section, which is approximately aligned so that the
$z$-coordinate is perpendicular to the disk plane, we place our imaginary
observer at $z=0$ and a solar radius of $8.2\,\kpc$ and vary the azimuth
between $0<\phi<2\pi$\footnote{Our very long-lived observer also has warp
drive.}. At each value of $\phi$ we then compute the median $z$ for stars
within a cylinder of radius $0.5\,\kpc$ and height $1\,\kpc$ perpendicular to
the fiducial disk. We then re-define this median $z$ as the new midplane of
the cylinder, re-select stars, and iterate until the median $z$ value
converges. We find that only $10$ iterations of this procedure are necessary
for convergence. The resulting median $z$ is taken to be what our observer
would measure as the local Galactic midplane at each $\phi$. This procedure
assumes perfect density estimation, and therefore perfect corrections for
extinction, within the cylinder defining the ``solar neighborhood.'' Imperfect
extinction correction is likely to increase the amplitude of the estimated
fluctuations in $z$. To account for the effect of particle noise, we
bootstrap-resample stars within a cylinder of stars with height $2\,\kpc$ and
the same radius $1000$ times and determine the $1\,\sigma$ error bars. This
bootstrap resampling is performed by first selecting all stars within a height
of $2\,\kpc$, resampling that selection, and then repeating the $10$
iterations.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=7in]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels). The local midplane is determined at a
position $\phi$ by taking the median height of all stars within $R=0.5\,\kpc$
and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma
$ error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=6.8in]{fig/range_dphi.pdf}
\end{center}
\caption{The range of midplane heights encountered as a function of angular
width. At each angle $\phi$ from Figure~\ref{fig:midplane} we consider an
angular width of $\Delta \phi$ centered on $\phi$ and report the range of
midplane heights within that width. We repeat the procedure for each $\phi$
and plot the result as translucent gray lines. We also plot the mean range as
a solid blue line and the $\pm1\sigma$ lines as dashed blue lines. The upper
$x$-axis shows the chord length from the position $\phi$ to $\pm\Delta\phi/2$.
The horizontal \thincolor{} and \thickcolor{} lines shows the midplane
variation necessary to induce a $10\%$ error in $J_z$ for the thin-disk and
thick-disk orbits, respectively. A cartoon explanation of this Figure is given
in Figure~\ref{fig:fig_to_explain}. We only consider the Solar circle to
explore the impact of azimuthal midplane variations, but similar effects are
expected for radial midplane variations.}
\label{fig:range_deltaphi}
\end{figure*}

@@

We also compute $R_{\n}$ for each of the orbits of the real open clusters in
Table~\ref{tab:real_clusters}, using the midplane variations from
\mi{} as a conservative choice, since it has the largest degree of
midplane variation (Figure~\ref{fig:midplane}). We report the neighborhood
values for the mass of each cluster in the neighborhood column of
Table~\ref{tab:real_clusters}. Except for alphaPer, we find that the
neighborhood around each cluster is typically $\gtrsim x\,\pc$. This analysis
assumes that the MW resembles \mi{}, which is not necessarily true. However,
this step of our analysis is meant to be more illustrative than quantitative:
this calculation should actually be performed using real measurements of
variations in the MW's stellar midplane. Such measurements should be possible
in the near future with {\em Gaia}, with quality distance estimates the
crucial limiting factor, along with careful selction function modelling.
